---
title: "Introduction to connectedness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to connectedness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What problem does this package solve?

In genetic evaluation, animals are often compared across herds, flocks,
regions, years, or other **management units (MUs)**. Those comparisons are not
equally reliable in all data sets. When management units are weakly linked by
pedigree or genomic relationships, differences in estimated breeding values
across units become less precise.

The **connectedness** package computes pairwise connectedness between MUs from
the same ingredients that define the evaluation context:

- the relationship structure among animals,
- the assignment of animals to management units,
- the fixed-effects design,
- and the assumed variance components.

The package supports connectedness computed from:

- **A-inverse** (pedigree-based connectedness),
- **G-inverse** (genomic connectedness),
- **H-inverse** (combined pedigree-genomic connectedness), and
- a **custom inverse kernel** supplied by the user.

## Which metrics does the package compute?

The package reports two complementary contrast-based metrics:

- **PEVD contrast**: prediction error variance of differences between MU
  contrasts. Lower values indicate stronger connectedness.
- **CD contrast**: coefficient of determination for the same contrasts. Higher
  values indicate stronger connectedness.

Both metrics are computed through mixed model equations (MME). The package uses
the relationship kernel and the fixed-effect structure to evaluate the precision
of pairwise MU contrasts, not a separate heuristic summary.

If desired, `PEVD` can be returned scaled by the additive genetic variance by
setting `scale_pevd = TRUE` in `compute_connectedness()`.

## What are target animals?

The **target animals** are the animals that receive non-zero weights in the
pairwise contrast vector. They are not necessarily the only animals used in the
mixed-model equations. Other animals may contribute information through their
records and relationships, but receive zero weight in a specific MU contrast.

This distinction is especially important when a time window is used. The MME can
be built using all records in `data`, while CD/PEVD are reported for contrasts
among a temporally defined set of target animals.

For a pair of MUs, target animals from one MU receive weights `1 / n_i`, target
animals from the other MU receive weights `-1 / n_j`, and all other animals
receive weight zero in that contrast. Those zero-weight animals can still affect
the solution through relationships and records in the MME.

## Quick start

For most users, the main entry point is `compute_connectedness()`.

A minimal pedigree-based analysis looks like this:

```{r eval=FALSE}
library(connectedness)

res <- compute_connectedness(
  data          = data,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree
)

print(res)
plot(res, which = "CD")
plot(res, which = "PEVD")
```

The short workflow is:

1. provide animals, MUs, and fixed effects,
2. choose the relationship representation,
3. obtain pairwise CD and PEVD between reported MUs.

## Inputs: what does the function need?

The package separates the problem into two layers.

### Evaluation context

These inputs define the comparison problem:

- `data`: records used to define animals, MUs, and fixed effects,
- `animal_col`: animal identifier column,
- `mu_col`: management-unit column,
- `fixed_formula`: fixed-effects structure,
- `sigma2a`, `sigma2e`: variance components.

### Relationship structure

These inputs define how animals are connected genetically:

- `relationship = "Ainv"` requires a pedigree,
- `relationship = "Ginv"` requires a genotype matrix `X` and an index linking
  row order to animal IDs,
- `relationship = "Hinv"` requires pedigree + genotypes + `genotyped_idx`,
- `relationship = "custom"` uses a supplied inverse kernel.

## Outputs: what do you get back?

`compute_connectedness()` returns an object of class `"connectedness"`. The most
important components are:

- `CD`: matrix of pairwise CD contrast values,
- `PEVD`: matrix of pairwise PEVD contrast values,
- `qK`: denominator of the contrast under the chosen kernel,
- `qC`: prediction error numerator of the contrast,
- `n_target`: number of target animals per reported MU,
- `report_mus`: MUs included in the reported CD/PEVD matrices,
- `target_scope`: definition of the target animals used in the contrasts,
- `relationship`: inverse relationship matrix type used,
- `activity_summary`: temporal MU activity summary when a time window and
  `min_records_per_year` are used,
- `overlap`: temporal overlap table when a time window is used.

When a time window is used, `n_target` refers to animals receiving non-zero
weights in the contrast. It does not necessarily include all animals used in the
MME.

If `scale_pevd = TRUE`, the `PEVD` matrix is returned on the scale
`PEVD / sigma2a`.

## Example: weak vs strong connectedness

The example below contrasts two small pedigree scenarios:

- a **weakly connected** setting, where the two MUs are linked only weakly,
- a **more connected** setting, where a sire is shared across MUs.

The goal is not biological realism, but to illustrate how the package responds
to a simple change in pedigree structure.

### Scenario 1: weaker connectedness

```{r eval=FALSE}
pedigree_weak <- data.frame(
  animal = c("S1", "S2", "D1", "D2", "A1", "A2", "B1", "B2"),
  sire   = c("0",  "0",  "0",  "0",  "S1", "S1", "S2", "S2"),
  dam    = c("0",  "0",  "0",  "0",  "D1", "D2", "D1", "D2")
)

data_weak <- data.frame(
  animal = c("A1", "A2", "B1", "B2"),
  MU     = c("MU1", "MU1", "MU2", "MU2"),
  sex    = c("M", "F", "M", "F")
)

res_weak <- compute_connectedness(
  data          = data_weak,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree_weak
)
```

### Scenario 2: stronger connectedness through a shared sire

```{r eval=FALSE}
pedigree_strong <- data.frame(
  animal = c("S1", "D1", "D2", "A1", "A2", "B1", "B2"),
  sire   = c("0",  "0",  "0",  "S1", "S1", "S1", "S1"),
  dam    = c("0",  "0",  "0",  "D1", "D2", "D1", "D2")
)

data_strong <- data.frame(
  animal = c("A1", "A2", "B1", "B2"),
  MU     = c("MU1", "MU1", "MU2", "MU2"),
  sex    = c("M", "F", "M", "F")
)

res_strong <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree_strong
)
```

### Compare the two results

```{r eval=FALSE}
res_weak$CD
res_weak$PEVD
res_weak$qK

res_strong$CD
res_strong$PEVD
res_strong$qK
```

The weak-versus-strong comparison is useful because PEVD and CD are related but
not identical summaries. A stronger pedigree link can reduce `PEVD`, improving
precision of the contrast. At the same time, the kernel-based denominator `qK`
can also change because the MUs become less genetically distinct. CD is
interpreted relative to that denominator.

## Optional PEVD scaling

```{r eval=FALSE}
res_strong_scaled <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree_strong,
  scale_pevd    = TRUE
)

res_strong_scaled$PEVD
```

## Relationship structures

A practical way to think about the four options is:

- use **Ainv** when the target notion of connectedness is pedigree-based,
- use **Ginv** when realized genomic links are the main object of interest,
- use **Hinv** when the connectedness measure should align with a combined
  pedigree-genomic evaluation context,
- use **custom** when your inverse kernel is built elsewhere.

### A-inverse

```{r eval=FALSE}
res_A <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree_strong
)
```

### G-inverse

```{r eval=FALSE}
X <- matrix(
  c(0, 1, 2, 1,
    1, 1, 2, 0,
    2, 1, 0, 1,
    1, 2, 1, 0),
  nrow = 4,
  byrow = TRUE
)
rownames(X) <- c("A1", "A2", "B1", "B2")

animal_index <- setNames(seq_len(nrow(X)), rownames(X))

res_G <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ginv",
  X             = X,
  animal_index  = animal_index
)
```

### H-inverse

```{r eval=FALSE}
renum <- renum_pedigree(pedigree_strong, verbose = FALSE)
genotyped_idx <- renum$new_id[match(rownames(X), renum$animal)]

res_H <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Hinv",
  pedigree      = pedigree_strong,
  X             = X,
  genotyped_idx = genotyped_idx
)
```

### Custom inverse kernels

Custom inverse kernels are useful when the relationship inverse is built by
external software or by a user-defined method. The key requirement is that row
and column identifiers match the names of `animal_index`.

The example below exports a package-built `Ainv` as triplets, reads it back, and
uses it through `relationship = "custom"`.

```{r eval=FALSE}
library(Matrix)

renum <- renum_pedigree(pedigree_strong, verbose = FALSE)
Ainv_res <- build_Ainv(renum)
Ainv_custom <- Ainv_res$Ainv

animal_index_custom <- stats::setNames(renum$new_id, renum$animal)
animal_index_custom <- animal_index_custom[!is.na(names(animal_index_custom))]

Ainv_triplet <- summary(Ainv_custom)
idx_to_id <- names(animal_index_custom)[order(animal_index_custom)]

Ainv_triplet_out <- data.frame(
  row = idx_to_id[Ainv_triplet$i],
  col = idx_to_id[Ainv_triplet$j],
  value = Ainv_triplet$x
)

write.table(
  Ainv_triplet_out,
  file = "Ainv_custom_triplet.txt",
  row.names = FALSE,
  col.names = TRUE,
  quote = FALSE
)

Ainv_triplet_in <- read.table(
  "Ainv_custom_triplet.txt",
  header = TRUE,
  stringsAsFactors = FALSE
)

i <- animal_index_custom[Ainv_triplet_in$row]
j <- animal_index_custom[Ainv_triplet_in$col]

Ainv_custom_in <- Matrix::sparseMatrix(
  i = i,
  j = j,
  x = Ainv_triplet_in$value,
  dims = c(length(animal_index_custom), length(animal_index_custom))
)

Ainv_custom_in <- Matrix::forceSymmetric(Ainv_custom_in, uplo = "L")
Ainv_custom_in <- methods::as(Ainv_custom_in, "dgCMatrix")

res_custom <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "custom",
  rel_matrix    = Ainv_custom_in,
  animal_index  = animal_index_custom
)

max(abs(res_A$CD - res_custom$CD), na.rm = TRUE)
max(abs(res_A$PEVD - res_custom$PEVD), na.rm = TRUE)
```

If a custom file stores only one triangle of a symmetric sparse matrix, the
matrix must be symmetrized before calling `compute_connectedness()`.

## Using a time window to define target animals

A time window is used to select MUs of interest and, by default, to define the
animals that receive non-zero weights in the contrast. The MME itself is still
built from all records supplied in `data`.

| `target_scope` | MME uses | Target animals | Question answered |
|---|---|---|---|
| `"window"` | all records in `data` | animals from selected MUs inside `year_window` | recent connectedness within the full system |
| `"selected_mus"` | all records in `data` | all animals from selected MUs | accumulated connectedness of active MUs |

```{r eval=FALSE}
pedigree_time <- data.frame(
  animal = c("S1", "D1", "D2", "A0", "A1", "A2", "B0", "B1", "B2"),
  sire   = c("0",  "0",  "0",  "S1", "S1", "S1", "S1", "S1", "S1"),
  dam    = c("0",  "0",  "0",  "D1", "D1", "D2", "D1", "D1", "D2")
)

data_time <- data.frame(
  animal = c("A0", "A1", "A2", "B0", "B1", "B2"),
  MU     = c("MU1", "MU1", "MU1", "MU2", "MU2", "MU2"),
  sex    = c("F", "M", "F", "F", "M", "F"),
  year   = c(2015, 2020, 2021, 2015, 2020, 2021)
)

res_window <- compute_connectedness(
  data                 = data_time,
  animal_col           = "animal",
  mu_col               = "MU",
  fixed_formula        = ~ 1 + sex,
  sigma2a              = 50,
  sigma2e              = 100,
  relationship         = "Ainv",
  pedigree             = pedigree_time,
  year_col             = "year",
  year_window          = c(2020, 2021),
  min_records_per_year = 1,
  target_scope         = "window"
)

res_selected_mus <- compute_connectedness(
  data                 = data_time,
  animal_col           = "animal",
  mu_col               = "MU",
  fixed_formula        = ~ 1 + sex,
  sigma2a              = 50,
  sigma2e              = 100,
  relationship         = "Ainv",
  pedigree             = pedigree_time,
  year_col             = "year",
  year_window          = c(2020, 2021),
  min_records_per_year = 1,
  target_scope         = "selected_mus"
)

res_window$n_target
res_selected_mus$n_target
```

With `target_scope = "window"`, the target animals are `A1`, `A2`, `B1`, and
`B2`. With `target_scope = "selected_mus"`, the target animals are `A0`, `A1`,
`A2`, `B0`, `B1`, and `B2`.

## Summarizing connectedness

The output matrices are pairwise. The helpers below summarize those pairwise
values; they do not create new contrasts against "all other MUs".

### Overall pairwise summaries

```{r eval=FALSE}
overall_connectedness <- function(res) {
  up <- upper.tri(res$CD)

  data.frame(
    n_mus = nrow(res$CD),
    n_pairs = sum(up),
    mean_CD = mean(res$CD[up], na.rm = TRUE),
    median_CD = stats::median(res$CD[up], na.rm = TRUE),
    mean_PEVD = mean(res$PEVD[up], na.rm = TRUE),
    median_PEVD = stats::median(res$PEVD[up], na.rm = TRUE)
  )
}

overall_connectedness(res_strong)
```

### MU-level summaries

```{r eval=FALSE}
mu_connectedness_summary <- function(res) {
  CD <- res$CD
  PEVD <- res$PEVD

  diag(CD) <- NA_real_
  diag(PEVD) <- NA_real_

  data.frame(
    MU = rownames(CD),
    mean_CD_to_others = rowMeans(CD, na.rm = TRUE),
    median_CD_to_others = apply(CD, 1, stats::median, na.rm = TRUE),
    mean_PEVD_to_others = rowMeans(PEVD, na.rm = TRUE),
    median_PEVD_to_others = apply(PEVD, 1, stats::median, na.rm = TRUE),
    n_target = res$n_target[rownames(CD)]
  )
}

mu_connectedness_summary(res_strong)
```

## Visualization

The package provides a direct plotting method for a first visual inspection.

```{r eval=FALSE}
plot(res_strong, which = "CD")
plot(res_strong, which = "PEVD")
plot(res_strong, which = "all")
plot(res_window, which = "overlap")
```

A quick visual check is often the easiest way to identify MU pairs that are well
connected, weakly connected, or effectively isolated.

## Computational notes

For large analyses, first inspect the problem size without solving:

```{r eval=FALSE}
diag <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree_strong,
  dry_run       = TRUE
)

diag$mme_dim
diag$schur_W_storage_mb
diag$selected_schur_solver
```

By default, `compute_connectedness()` uses the Schur-complement MME solving
method (`mme_backend = "schur"`). The previous full-MME method remains available
for validation in small data sets with `mme_backend = "full_mme"`.

With `schur_solver = "auto"`, sparse inverse kernels such as `Ainv` are routed
through CHOLMOD via the Matrix package, whereas dense kernels such as `Ginv` are
routed to a dense compiled solver. If the full dense Schur working matrix
`Cuu^{-1} Z'X` would be too large, `"auto"` selects `"cholmod_lowmem"`, which
constructs the same fixed-effect Schur complement in blocks and solves
`Cuu^{-1} Z'X b` on demand during MU block processing. The `"eigen_sparse"`
option is mainly a diagnostic path for small comparisons.

## References

Kennedy, B. W., & Trus, D. (1993). Considerations on genetic connectedness
between management units under an animal model. *Journal of Animal Science*,
71, 2341–2352.

Laloë, D. (1993). Precision and information in linear models of genetic
evaluation. *Genetics Selection Evolution*, 25, 557–576.

Laloë, D., Phocas, F., & Ménissier, F. (1996). Considerations on measures of
precision and connectedness in mixed linear models of genetic evaluation.
*Genetics Selection Evolution*, 28, 359–378.

Yu, H., & Morota, G. (2021). GCA: An R package for genetic connectedness
analysis using pedigree and genomic data. *BMC Genomics*, 22, 119.
