---
title: "Introduction to connectedness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to connectedness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What problem does this package solve?

In genetic evaluation, animals are often compared across herds, flocks,
regions, years, or other **management units (MUs)**. Those comparisons are not
equally reliable in all data sets. When management units are weakly linked by
pedigree or genomic relationships, differences in estimated breeding values
across units become less precise.

The **connectedness** package addresses that problem. It computes pairwise
connectedness between MUs from the same ingredients that define the evaluation
context:

- the relationship structure among animals,
- the assignment of animals to management units,
- the fixed-effects design,
- and the assumed variance components.

The package currently supports connectedness computed from:

- **A-inverse** (pedigree-based connectedness),
- **G-inverse** (genomic connectedness),
- **H-inverse** (single-step connectedness), and
- a **custom inverse kernel** supplied by the user.

## Which metrics does the package compute?

The package reports two complementary metrics:

- **PEVD contrast**: prediction error variance of differences between MU
  contrasts. Lower values indicate stronger connectedness.
- **CD contrast**: coefficient of determination for the same contrasts. Higher
  values indicate stronger connectedness.

Both metrics are computed under the **contrast approach** through mixed model
equations. That point is important: the package is not combining unrelated
metrics, but two connected summaries derived from the same contrast framework.
This is one of the reasons why the package can also exploit ideas such as
**temporal overlap** between units when records are restricted to a common time
window.

## Quick start

For most users, the main entry point is `compute_connectedness()`.

A minimal pedigree-based analysis looks like this:

```{r eval=FALSE}
library(connectedness)

res <- compute_connectedness(
  data          = data,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree
)

print(res)
plot(res, which = "CD")
plot(res, which = "PEVD")
```

That is the shortest mental model of the package:

1. provide the animals, MUs, and fixed-effect structure,
2. choose the relationship representation,
3. obtain pairwise CD and PEVD between MUs.

## Inputs: what does the function need?

The package separates the problem into two layers.

### 1. Evaluation context

These inputs define *what comparison problem* you want to study:

- `data`: records used to define the animals and management units involved,
- `animal_col`: animal identifier column,
- `mu_col`: management unit column,
- `fixed_formula`: fixed-effects structure,
- `sigma2a`, `sigma2e`: variance components.

### 2. Relationship structure

These inputs define *how animals are connected genetically*:

- `relationship = "Ainv"` requires a pedigree,
- `relationship = "Ginv"` requires a genotype matrix `X` and an index linking
  row order to animal IDs,
- `relationship = "Hinv"` requires pedigree + genotypes + `genotyped_idx`,
- `relationship = "custom"` uses a supplied inverse kernel.

So the workflow is intentionally explicit: the user chooses not only the data,
but also the **type of connectedness** they want to quantify.

## Outputs: what do you get back?

`compute_connectedness()` returns an object of class `"connectedness"`.
The most important components are:

- `CD`: matrix of pairwise CD contrast values,
- `PEVD`: matrix of pairwise PEVD contrast values,
- `qK`: denominator of the contrast under the chosen kernel,
- `qC`: prediction error numerator of the contrast,
- `n_target`: number of target animals per MU,
- `relationship`: the inverse relationship matrix used,
- `overlap`: temporal overlap table when a time window is used.

## Example

A good introductory vignette should let the user see the package react to a
simple and meaningful change in connectedness. The example below contrasts two
small pedigree scenarios:

- a **weakly connected** setting, where the two MUs are linked only weakly,
- a **more connected** setting, where a sire is shared across MUs.

The goal is not biological realism, but to show immediately that the metrics
move in the expected direction.

### Scenario 1: weaker connectedness

```{r eval=FALSE}
pedigree_weak <- data.frame(
  animal = c("S1", "S2", "D1", "D2", "A1", "A2", "B1", "B2"),
  sire   = c("0",  "0",  "0",  "0",  "S1", "S1", "S2", "S2"),
  dam    = c("0",  "0",  "0",  "0",  "D1", "D2", "D1", "D2")
)

data_weak <- data.frame(
  animal = c("A1", "A2", "B1", "B2"),
  MU     = c("MU1", "MU1", "MU2", "MU2"),
  sex    = c("M", "F", "M", "F")
)

res_weak <- compute_connectedness(
  data          = data_weak,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree_weak
)
```

### Scenario 2: stronger connectedness through a shared sire

```{r eval=FALSE}
pedigree_strong <- data.frame(
  animal = c("S1", "D1", "D2", "A1", "A2", "B1", "B2"),
  sire   = c("0",  "0",  "0",  "S1", "S1", "S1", "S1"),
  dam    = c("0",  "0",  "0",  "D1", "D2", "D1", "D2")
)

data_strong <- data.frame(
  animal = c("A1", "A2", "B1", "B2"),
  MU     = c("MU1", "MU1", "MU2", "MU2"),
  sex    = c("M", "F", "M", "F")
)

res_strong <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree_strong
)
```

### Compare the two results

```{r eval=FALSE}
res_weak$CD
res_weak$PEVD

res_strong$CD
res_strong$PEVD
```

The expected pattern is straightforward:

- the **more connected** scenario should tend to show **higher CD**,
- and **lower PEVD**,
- although the exact numerical behavior depends on the full contrast structure
  and the implied genetic denominator.

This is the kind of first experience a new user needs: a small example where it
is obvious that the package is responding to the pedigree structure rather than
producing opaque numbers.

## Choosing among Ainv, Ginv, Hinv, and custom

A practical way to think about the four options is:

- use **Ainv** when the target notion of connectedness is pedigree-based,
- use **Ginv** when realized genomic links are the main object of interest,
- use **Hinv** when you want the connectedness measure to align with a
  single-step evaluation context,
- use **custom** when your inverse kernel is built elsewhere.

### Example: G-inverse

```{r eval=FALSE}
X <- matrix(
  c(0, 1, 2, 1,
    1, 1, 2, 0,
    2, 1, 0, 1,
    1, 2, 1, 0),
  nrow = 4,
  byrow = TRUE
)
rownames(X) <- c("A1", "A2", "B1", "B2")

animal_index <- setNames(seq_len(nrow(X)), rownames(X))

res_G <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ginv",
  X             = X,
  animal_index  = animal_index
)
```

### Example: H-inverse

```{r eval=FALSE}
renum <- renum_pedigree(pedigree_strong, verbose = FALSE)
genotyped_idx <- renum$new_id[match(rownames(X), renum$animal)]

res_H <- compute_connectedness(
  data          = data_strong,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Hinv",
  pedigree      = pedigree_strong,
  X             = X,
  genotyped_idx = genotyped_idx
)
```

## Temporal overlap: why the contrast framework helps

The package can optionally restrict the analysis to a common time window and
report where MU pairs actually overlap in time.

```{r eval=FALSE}
data_time <- transform(
  data_strong,
  year = c(2020, 2021, 2020, 2021)
)

res_time <- compute_connectedness(
  data                 = data_time,
  animal_col           = "animal",
  mu_col               = "MU",
  fixed_formula        = ~ 1 + sex,
  sigma2a              = 50,
  sigma2e              = 100,
  relationship         = "Ainv",
  pedigree             = pedigree_strong,
  year_col             = "year",
  year_window          = c(2020, 2021),
  min_records_per_year = 1
)

plot(res_time, which = "overlap")
```

This is a useful extension of the contrast logic: connectedness is not only a
function of pedigree or genomics, but also of **which units are effectively
represented in the same time horizon**.

## Visualization

The package provides a direct plotting method for a first visual inspection.

```{r eval=FALSE}
plot(res_strong, which = "CD")
plot(res_strong, which = "PEVD")
plot(res_strong, which = "all")
```

A quick visual check is often the easiest way to identify MU pairs that are
well connected, weakly connected, or effectively isolated.

## References

Kennedy, B. W., & Trus, D. (1993). Considerations on genetic connectedness
between management units under an animal model. *Journal of Animal Science*,
71, 2341–2352. fileciteturn55file0

Laloë, D. (1993). Precision and information in linear models of genetic
evaluation. *Genetics Selection Evolution*, 25, 557–576.

Laloë, D., Phocas, F., & Ménissier, F. (1996). Considerations on measures of
precision and connectedness in mixed linear models of genetic evaluation.
*Genetics Selection Evolution*, 28, 359–378.
