---
title: "Introduction to connectedness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to connectedness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why connectedness matters

When animals are compared across herds, flocks, regions, or other management
units (MUs), the reliability of those comparisons depends on the degree of
*genetic connectedness* among units. Under an animal model, connectedness is
not only a consequence of shared sires: it arises through the full relationship
structure among animals.

A central practical implication is that comparisons between animals raised in
different MUs become less reliable when the units are weakly connected. In the
animal-model setting, Kennedy and Trus (1993) argued that the most meaningful
measure of connectedness is the prediction error variance of differences in EBV
between animals from different MUs, because it directly quantifies the error of
the comparison that matters for selection. They also showed that genetic links
*across* MUs reduce prediction error, whereas strong relationships concentrated
*within* MUs can actually worsen across-unit comparisons by increasing drift and
the uncertainty of MU contrasts. Kennedy and Trus (1993) also evaluated simpler
surrogate measures and found that quantities tied to the variance of estimated
MU differences were much more informative than crude gene-flow summaries. fileciteturn55file0

The **connectedness** package implements this logic through two complementary
metrics computed from mixed model equations:

- **CD**: Coefficient of Determination of the contrast between MU means.
  Higher values indicate stronger connectedness.
- **PEVD**: Prediction Error Variance of Differences.
  Lower values indicate stronger connectedness.

The package can now compute connectedness from:

- a pedigree-based inverse relationship matrix (**A-inverse**),
- a genomic inverse relationship matrix (**G-inverse**),
- a single-step inverse relationship matrix (**H-inverse**), or
- a user-supplied inverse kernel.

## What this package does

The core workflow has two steps:

1. build or provide an inverse relationship matrix,
2. compute CD and PEVD between all pairs of management units.

In most use cases, you will call `compute_connectedness()` directly and choose
which relationship structure should underlie the analysis.

## Minimal data objects

You typically need at least:

- `data`: phenotypic records including animal ID, management unit, and any
  fixed effects required in the model matrix,
- `pedigree`: a three-column pedigree (animal, sire, dam) when using pedigree
  or single-step connectedness,
- `X`: a genotype matrix when using genomic or single-step connectedness.

```{r eval=FALSE}
library(connectedness)

pedigree <- data.frame(
  animal = c("S1", "S2", "D1", "D2", "A", "B", "C", "D"),
  sire   = c("0",  "0",  "0",  "0",  "S1", "S1", "S2", "S2"),
  dam    = c("0",  "0",  "0",  "0",  "D1", "D2", "D1", "D2")
)

data <- data.frame(
  animal     = c("A", "B", "C", "D"),
  MU         = c("North", "North", "South", "South"),
  birth_year = c(2020, 2020, 2020, 2020),
  sex        = c("M", "F", "M", "F"),
  weight     = c(320, 310, 340, 330)
)
```

## Typical workflow with A-inverse

This is the default pedigree-based analysis. The package renumbers the pedigree,
builds `Ainv`, and computes connectedness from the mixed model equations.

```{r eval=FALSE}
res_A <- compute_connectedness(
  data          = data,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ainv",
  pedigree      = pedigree
)

print(res_A)
plot(res_A, which = "CD")
plot(res_A, which = "PEVD")
```

## Using G-inverse

When a genomic relationship structure is preferred, provide a genotype matrix
`X` coded as 0/1/2. In this case connectedness is computed from `Ginv`.

```{r eval=FALSE}
# toy genotype matrix: rows must correspond to the genotyped animals
X <- matrix(
  c(0, 1, 2, 1,
    1, 1, 2, 0,
    2, 1, 0, 1,
    1, 2, 1, 0),
  nrow = 4,
  byrow = TRUE
)
rownames(X) <- c("A", "B", "C", "D")

animal_index <- setNames(seq_len(nrow(X)), rownames(X))

res_G <- compute_connectedness(
  data          = data,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Ginv",
  X             = X,
  animal_index  = animal_index
)
```

This is appropriate when you want connectedness to reflect realized genomic
relationships among the genotyped animals rather than pedigree expectation.

## Using H-inverse

Single-step connectedness combines pedigree and genomic information. Here the
rows of `X` must correspond to the pedigree indices given by `genotyped_idx`.

```{r eval=FALSE}
renum <- renum_pedigree(pedigree, verbose = FALSE)

genotyped_idx <- renum$new_id[match(c("A", "B", "C", "D"), renum$animal)]

res_H <- compute_connectedness(
  data          = data,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "Hinv",
  pedigree      = pedigree,
  X             = X,
  genotyped_idx = genotyped_idx
)
```

This option is useful when some animals are genotyped and you want connectedness
to reflect the same single-step structure used in genomic evaluation.

## Using a custom inverse kernel

Advanced users can supply their own inverse kernel directly.

```{r eval=FALSE}
res_custom <- compute_connectedness(
  data          = data,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  relationship  = "custom",
  rel_matrix    = my_Kinv,
  animal_index  = my_index
)
```

This is the most flexible route when the relationship structure is built
outside the package.

## Temporal restrictions

In many data sets, MUs are not connected uniformly through time. The package can
restrict the analysis to a given time window and report the years in which MU
pairs overlap.

```{r eval=FALSE}
res_window <- compute_connectedness(
  data                 = data,
  animal_col           = "animal",
  mu_col               = "MU",
  fixed_formula        = ~ 1 + sex,
  sigma2a              = 50,
  sigma2e              = 100,
  relationship         = "Ainv",
  pedigree             = pedigree,
  year_col             = "birth_year",
  year_window          = c(2018, 2022),
  min_records_per_year = 5
)

plot(res_window, which = "overlap")
```

## Interpreting CD and PEVD

CD and PEVD should be interpreted jointly:

- **higher CD** means more precise contrasts between MUs,
- **lower PEVD** means less uncertainty in EBV differences across MUs.

For many practical applications, PEVD is the most directly interpretable metric
because it quantifies the prediction error of the comparison itself. This is
precisely why Kennedy and Trus (1993) emphasized prediction error of differences
as the most relevant connectedness criterion under an animal model. fileciteturn55file0

The package therefore exposes both views: `CD` as a scaled connectedness metric,
and `PEVD` as the underlying error scale.

## Visualization

The `plot()` method can display heatmaps for CD and PEVD and, when available,
a temporal overlap plot.

```{r eval=FALSE}
plot(res_A, which = "CD")
plot(res_A, which = "PEVD")
plot(res_A, which = "all")

# optional controls
plot(res_A, which = "CD", show_values = TRUE, triangle = "upper")
```

## Building inverse matrices explicitly

Although `compute_connectedness()` is the main entry point, the package also
provides wrappers to build inverse relationship matrices directly.

```{r eval=FALSE}
renum <- renum_pedigree(pedigree, verbose = FALSE)
Ainv_res <- build_Ainv(renum)
Ginv_res <- build_Ginv(X)
Hinv_res <- build_Hinv(renum, genotyped_idx = genotyped_idx, X = X)
```

These functions are useful when you want to inspect intermediate matrices or
reuse them across analyses.

## Suggested vignette strategy

At the current stage of the package, a **single strong introductory vignette**
is sufficient, provided that it clearly explains:

- why connectedness matters,
- how to choose among `Ainv`, `Ginv`, `Hinv`, and `custom`,
- how to interpret CD and PEVD,
- and how to visualize the results.

That said, once the package stabilizes, a second vignette would likely add real
value. A good split would be:

1. **Introductory vignette**: concepts, main workflow, interpretation.
2. **Advanced vignette**: genomic and single-step workflows, tuning of `G`,
   temporal analyses, and larger-scale use cases.

My view is that right now **one vignette is enough**, but only if it is much
more purposeful than the current version.

## References

Kennedy, B. W., & Trus, D. (1993). Considerations on genetic connectedness
between management units under an animal model. *Journal of Animal Science*,
71, 2341–2352. fileciteturn55file0

Laloë, D. (1993). Precision and information in linear models of genetic
evaluation. *Genetics Selection Evolution*, 25, 557–576.

Laloë, D., Phocas, F., & Ménissier, F. (1996). Considerations on measures of
precision and connectedness in mixed linear models of genetic evaluation.
*Genetics Selection Evolution*, 28, 359–378.
