---
title: "Introduction to connectedness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to connectedness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The **connectedness** package computes two complementary metrics of
connectedness between management units (MUs) in animal breeding evaluations:

- **CD**: Coefficient of Determination of the contrast between MU means.  
  Ranges from 0 (no connectedness) to 1 (perfect connectedness).
- **PEVD**: Prediction Error Variance of Differences between MU means.  
  In the same units as the residual variance ($\sigma^2_e$). Lower is better.

Both metrics are based on the contrast approach via Mixed Model Equations
(Laloë, 1993; Laloë et al., 1996).

## Typical workflow

### 1. Prepare your data

You need two objects:

- `pedigree`: a data frame with three columns (animal, sire, dam).  
  Should include all ancestors, even those without phenotypic records.  
  Unknown parents should be coded as `"0"`.

- `data`: a data frame with phenotypic records, including columns for  
  animal ID, MU membership, and fixed effects.

```{r eval=FALSE}
library(connectedness)

# Example pedigree (complete, including ancestors)
pedigree <- data.frame(
  animal = c("S1","S2","D1","D2","A","B","C","D"),
  sire   = c("0","0","0","0","S1","S1","S2","S2"),
  dam    = c("0","0","0","0","D1","D2","D1","D2")
)

# Example phenotypic data
data <- data.frame(
  animal     = c("A","B","C","D","A","C"),
  MU         = c("North","North","South","South","North","South"),
  birth_year = c(2010,2010,2011,2011,2012,2012),
  sex        = c("M","F","M","F","M","F"),
  weight     = c(320,310,340,330,325,335)
)
```

### 2. Run the analysis

```{r eval=FALSE}
res <- compute_connectedness(
  data          = data,
  pedigree      = pedigree,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100
)
```

The variance components (`sigma2a`, `sigma2e`) should come from a prior
REML or Bayesian variance components analysis of your data.

### 3. Inspect results

```{r eval=FALSE}
print(res)
#> Connectedness analysis
#> ...
#> Coefficient of Determination (CD):
#>        North  South
#> North     NA  0.412
#> South  0.412     NA
#>
#> PEV of Differences (PEVD):
#>        North  South
#> North     NA 58.231
#> South 58.231     NA

plot(res, which = "CD")
```

### 4. Using a temporal window

If your data spans many years, you may want to restrict the analysis to
animals with records in a common period:

```{r eval=FALSE}
res_window <- compute_connectedness(
  data                 = data,
  pedigree             = pedigree,
  animal_col           = "animal",
  mu_col               = "MU",
  fixed_formula        = ~ 1 + sex,
  sigma2a              = 50,
  sigma2e              = 100,
  year_col             = "birth_year",
  year_window          = c(2010, 2022),
  min_records_per_year = 10
)

# visualize temporal overlap between MU pairs
plot(res_window, which = "overlap")
```

### 5. Using a custom relationship matrix (genomic)

If you have a pre-built inverse relationship matrix (e.g. G-inverse from
genomic data), you can pass it directly:

```{r eval=FALSE}
# animal_index maps original IDs to row/column positions in rel_matrix
animal_index <- setNames(1:4, c("A","B","C","D"))

res_g <- compute_connectedness(
  data          = data,
  animal_col    = "animal",
  mu_col        = "MU",
  fixed_formula = ~ 1 + sex,
  sigma2a       = 50,
  sigma2e       = 100,
  rel_matrix    = my_Ginv,       # dgCMatrix, N x N
  animal_index  = animal_index
)
```

**Important**: when using `rel_matrix` directly, you are responsible for
ensuring that (1) the matrix is the *inverse* of the relationship matrix,
scaled consistently with `sigma2a`; (2) `data` contains only animals present
in `animal_index`; and (3) the matrix is symmetric and positive definite.

## References

Laloë, D. (1993). Precision and information in linear models of genetic
evaluation. *Genetics Selection Evolution*, 25, 557–576.

Laloë, D., Phocas, F., & Ménissier, F. (1996). Considerations on measures
of precision and connectedness in mixed linear models of genetic evaluation.
*Genetics Selection Evolution*, 28, 359–378.
